Classless Java: Tuning Java Interfaces

### 2015-11-26 paper/sections/Overview.tex
Overview

\paragraph{Summary} Dealing with state (whether mutable or immutable)
in object interfaces relies on a notion of abstract state, where only
methods to interact with state are available to users. Object
interfaces provides support for four different types abstract state
operations, which are summarized in Figure~\ref{fig:abstractstate}.
The abstract state operations are determined by naming conventions and the
types of the methods. Fluent setters are a variant of mutable setters,
and are discussed in the case study in Section~\ref{}.

@@

\paragraph{Field type refinement with object interfaces}
In our proposed solution,
\begin{itemize}
\item by covariant method overriding we refine the field type
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}

@@

From the type perspective, the key is the covariant method overriding; however from the semantic perspective the key is the implementation for the setter with the old signature:
by using the \Q@with@ method we may use the information for \Q@z@ already stored in the object to forge an appropriate \Q@Point3D@ to store.
Note how all the informations about what fields sits in \Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the \Q@with@ method, and is transparent for the implementer of \Q@Bird@.

\bruno{See how to fit this in.}
\subsubsection*{\Q@Point3D@ and properties updater}
\Q@Bird@ needs a 3D location, thus first of all we define \Q@Point3D@ and
as follows:
\lstinputlisting[linerange=51-55]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\marco{Fix over: with(Point must be with(Point2D
and it should be after withZ, and on more lines}
Just to give you a feeling on how much boring code \mixin is generating,  we show the expanded code without then \mixin annotation\footnote{
As you can see, the generated code is very repetitive, writing such code by hand can easily induce bugs, for example a distract programmer may
swap the arguments of one of the many calls of \Q@Point3D.of@.
}.\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);    Point3D withY(int val);    Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int _x, int _y, int _z){
    int x=_x;    int y=_y;    int z=_z;
    return new Point3D(){
      public int x(){return x;}    public int y(){return y;}    public int z(){return z;}
      public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
      public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
      public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
      public Point3D with(Point2D val){
        if(val instanceof Point3D){return (Point3D)val;}
        return Point3D.of(val.x(),val.y(),this.z());
        }      };    }  }
\end{lstlisting}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin do not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

\subsection{Advanced Multiple Inheritance with Object Interfaces}
The \emph{``multiple inheritance''} case appears at interface
\texttt{Pegasus}.
Pegasus (one of the best known creatures
in Greek mythology) can not only \emph{run} but also \emph{fly}! 
It would be handy if they could obtain \texttt{fly} and \texttt{run} functionality from both \texttt{Horse} and \texttt{Bird}.

 Using interfaces and \mixin annotation,
this can happen transparently.

@@

Note how  even the non trivial pattern for field type refinement is transparently composed, and our pegasus have a \Q@Point3D@ \Q@location@.
Note that this works because \Q@Horse@ do not perform any field type refinement , otherwise we may have to choose/create a common subtype in order for pegasus to exists.

>>>>

\paragraph{Summary} Dealing with state (whether mutable or immutable)
in object interfaces relies on a notion of abstract state, where only
methods to interact with state are available to users. Object
interfaces provides support for four different types abstract state
operations, which are summarized in Figure~\ref{fig:abstractstate}.
The abstract state operations are determined by naming conventions and the
types of the methods. Fluent setters are a variant of mutable setters,
and are discussed in the case study in Section~\ref{sec:dsls}.

@@

\paragraph{Field type refinement with object interfaces}
Object interfaces address the challange of type-refinement as follows:
\begin{itemize}
\item by covariant method overriding we refine the type of the location field to \texttt{Point3D};
\item by \emph{overloading} we define a new setter for the location field with the more precise type;
\item using a default method we provide an implementation for the setter called 
with the old signature.
\end{itemize}

@@

\noindent From the type perspective, the key is the covariant method
overriding of \texttt{location()}. However from the semantic
perspective the key is the implementation for the setter with the old
signature.

\paragraph{\Q@Point3D@ and properties updater}
To implement the old setter in a convenient way, \mixin supports one
last type of operations: property updater \texttt{with}
methods. Unlike the \texttt{withX} (where \texttt{X} stands for a
field name) methods presented so far, property updaters take several
fields at once, contained in an interface, and copy those fields into
fields of another interface.
The \Q@Point3D@ interface is defined as follows:
\lstinputlisting[linerange=51-55]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D
\noindent By using the new \Q@with@ method we may use the information for
\Q@z@ already stored in the object to forge an appropriate \Q@Point3D@
to store.  Note how all the informations about what fields sits in
\Q@Point3D@ and what in \Q@Point2D@ is properly encapsulated in the
\Q@with@ method, and is transparent for the implementer of \Q@Bird@.

\paragraph{Generated Boilerplate}
Just to give a feeling on how much boring code \mixin is generating,
we show the generated code for the \texttt{Point3D} interface 
Figure~\ref{fig:boilerplate}. The generated code is very repetitive,
writing such code by hand can easily induce bugs, for example a
distracted programmer may swap the arguments of one of the many calls of
\Q@Point3D.of@.
\begin{figure}
\begin{lstlisting}
interface Point3D extends Point2D{
  Point3D withX(int val);    Point3D withY(int val);    Point3D withZ(int val);
  Point3D with(Point2D val);
  public static Point3D of(int _x, int _y, int _z){
    int x=_x;    int y=_y;    int z=_z;
    return new Point3D(){
      public int x(){return x;}    public int y(){return y;}    public int z(){return z;}
      public Point3D withX(int val){return Point3D.of(val,this.y(),this.z());}
      public Point3D withY(int val){return Point3D.of(this.x(),val,this.z());}
      public Point3D withZ(int val){return Point3D.of(this.x(),this.y(),val);}
      public Point3D with(Point2D val){
        if(val instanceof Point3D){return (Point3D)val;}
        return Point3D.of(val.x(),val.y(),this.z());
        }      };    }  }
\end{lstlisting}
\caption{Generated boilerplate code.}
\label{fig:boilerplate}
\end{figure}
Note how \Q@with-@ methods are automatically refined in their return type, so that code like 
\Q@Point3D p=Point3D.of(1,2,3); p=p.withX(42);@ will be accepted and behave as expected. If the programmer wishes to suppress this behaviour and keep the signature as it was, it is sufficient to redefine the \Q@with-@ methods in the new class repeating the old signature.
Again, the philosophy is that if the programmer provides something directly, \mixin does not touch it.
\Q@with-@ methods (functionally) update a field/property at a time.
This can be inefficient, and sometime hard to maintain.
Often we want to update many fields at the same time, for example using another object as source.
Following this idea, the method \Q@with(Point2D)@ is an example of a (functional) properties updater: 
it takes a certain type and in the current object update all field in that type that match fields in the current type.
The idea is that we want as result something that is still like \Q@this@, but modified to be as much as possible similar to the parameter.
The cast in \Q@with(Point2D)@ is trivially safe since it is guarded by an \Q@instanceof@ test. The idea is that if the parameter is a subtype of the current exact type, then we can just return the parameter, as something that is just ``more'' than \Q@this@.

\subsection{Advanced Multiple Inheritance with Object Interfaces}
Finally, we can define Pegasus as simply as we did in the simplified (and stateless) 
version in the right-side of Figure~\ref{fig:pegasus}:

@@

\noindent Note how even the non trivial pattern for field type refinement is
transparently composed, and \texttt{Pegasus} has a \Q@Point3D@
\Q@location@. This works because \Q@Horse@ do not perform
any field type refinement, otherwise we may have to choose/create a
common subtype in order for \texttt{Pegasus} to exists.
